### 2016-07-20 Projects/DelCo_2012/Report/DelCo_2012.tex
More work on GC discussion

Agents designed primarily for Class A (ordinary combustible) fires have been used most extensively in conjunction with wildland or wildland urban interface fires. These agents are said to be more effective than plain water while being environmentally safe. In some cases, they are also claimed to reduce the quantity and toxicity of smoke.

So the question is, does CAF provide improved gas cooling when compared to plan water?

The gas cooling experiments were designed to measure the ability of water streams and CAF streams to cool a hot gas layer with a minimum temperature of 250~$^{\circ}$C (482~$^{\circ}$F) at positions 1.83~m (6~ft) below the ceiling. During the planning of the experiments, some workshop attendees thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAFS, since the injected air, air being a great insulator, would interfere and result in less cooling. It was the belief that the reduced gas cooling capabilities of CAF may have led to firefighter injuries during interior fire attacks. Other firefighters thought that the bubble structure would provide the same mass flowrate of water with more surface area, which would enable a faster heat transfer rate and therefore, improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling.

In each gas cooling experiment, the same mass flow rate of water was used for both the plain water stream and the CAF stream. Typically the volumetric flow rate of water was 7.57~L/s (120~gpm). The density of water is 1000~kg/m$^3$ (8.3~lb/gal), therefore the mass flow rate was 7.57~kg/s (16.7~lb/s). The average time of flow was 15~s, therefore a total of 114~kg (251~lb) of water was introduced into each gas cooling experiment.

When a CAF stream was used it contained 0.3\% foam concentration by volume, the density of the foam concentrate is 1055~kg/m$^3$ (8.8~lb/gal), for a 15~s flow time, 3.6~kg (8.0 lb) of concentrate was used. In addition to the foam concentrate mixed with the water in the CAF stream is compressed air. The volume of air injected into the foam solution is 0.03~m$^3$/s (60~cfm). The density of the compressed air in the range of 100 to 150~psig, would be between 8.9~kg/m$^3$ (0.6 lb/ft$^3$) and 12.8~kg/m$^3$ (0.8 lb/ft$^3$). So a 15~s CAF flow would contain approximately 4.0~kg (8.8 lb) to 9.5~kg (21.0 lb) of air. Comparing the mass flows, and using the air mass based on the higher pressure, water still makes up more than 90~\% of the mass of the CAF stream.

Under the given conditions of a residential scale room and the mass of water introduced into the hot gas layer, it should be no surprise that the water is doing the work. Air filled bubbles in the CAFS stream had no discernible impact in these experiments.
At the same time there was no measurable negative effect of the air bubbles with regard to gas cooling.

>>>>

While CAF has been readily accepted for use in the wildland and wildland urban interface, the effectivenss of CAF for interior structural firefighting varies widely.  For example, does CAF provide improved gas cooling when compared to plan water?  During the planning of the experiments, some technical panel members thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAFS.  Their position was based on the thought that CAF has injected air.  Since air is a great insulator, it slow heat transfer and result in less cooling. It was their belief that the reduced gas cooling capabilities of CAF may have led to firefighter injuries during interior fire attacks. Other member of the technical panel thought that the bubble structure would provide the same mass flowrate of water but with more surface area, which would enable a faster heat transfer rate and therefore, improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling.

The gas cooling experiments were designed to measure the ability of water streams and CAF streams to cool a hot gas layer, without any application of agent to the source of the heat.  The 3.35~m (11~ft) tall ceiling in the burn building allowed a hot gas layer,at least 1.83~m (6 ft) to form. During the testing, a minimum temperature of 250~$^{\circ}$C (482~$^{\circ}$F) at positions 1.83~m (6~ft) below the ceiling was attained prior to the beginning of another gas cooling experiment.

In each gas cooling experiment, the same mass flow rate of water was used for both the plain water stream and the CAF stream. Typically the volumetric flow rate of water was 7.57~L/s (120~gpm). The density of water is 1000~kg/m$^3$ (8.3~lb/gal), therefore the mass flow rate was 7.57~kg/s (16.7~lb/s). The average time of flow was approximately 15~s, therefore a total of 114~kg (251~lb) of water was introduced into each gas cooling experiment.

When a CAF stream was used it contained 0.3\% foam concentration by volume, the density of the foam concentrate is 1055~kg/m$^3$ (8.8~lb/gal), for a 15~s flow time, 0.34~kg (0.75~lb) of concentrate was used. In addition to the foam concentrate and water, the CAF stream has compressed air. The volume of air injected into the foam solution is 0.03~m$^3$/s (60~cfm). The density of the compressed air in the range of 100 to 150~psig, would be between 8.9~kg/m$^3$ (0.6 lb/ft$^3$) and 12.8~kg/m$^3$ (0.8 lb/ft$^3$). So a 15~s CAF flow would contain approximately 4.0~kg (8.8 lb) to 9.5~kg (21.0 lb) of air. Comparing the mass flows, and using the air mass based on the higher pressure, water still makes up more than 90~\% of the mass of the CAF stream.

The estimated total expanded uncertainty of the temperature measurements in these experiments is $\pm$ 15~\%.  In other words, for a difference between one measurement and another measurement to be considered significantly different, the difference would have to be greater by more than $\pm$ 15~\%.  If the difference between two measurements was $\pm$ 15~\% or less, the results are considered to be similar.

As pointed out in the results section, the conditions at TC array 3 represented the "best" gas cooling results with the least impact of agents hitting and cooling the thermocouples themselves.

Under the given conditions of a residential scale room and the amount of water introduced into the hot gas layer, it should be no surprise that the water is doing the work. Air filled bubbles in the CAFS stream had no discernible impact, either positive or negative, in the gas cooling experiments.
At the same time there was no measurable negative effect of the air bubbles with regard to gas cooling.
